Incremental Forward Reasoning for White-Box~Proof~Search

### 2025-02-17 paper.tex
Correct and expand PVS info

% TODO: standardise indices vs indexes
% TODO: standardise dots after equations
% TODO: rename variable maps to variable indices everywhere
% TODO: standardise Rocq vs Coq

@@

In many proof assistants, simple automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's (TODO ref) and Rocq's (TODO ref) auto, ACL2's (TODO ref) waterfall, PVS's (TODO ref) grind, Lean's (TODO ref) Aesop (TODO ref), etc.
These tactics operate on \emph{goals} $Γ ⊢ A$, where the \emph{context} $Γ$ is a list of assumptions and $A$ is the proposition to be proved.
Faced with an initial goal, the tactics try various user-specified \emph{rules}, which correspond to admissible inferences in the prover's logic.
A successful rule reduces the current goal to zero or more subgoals, which are solved recursively until either a full proof is found or no rule applies to any open goal or the search times out.

@@

\paragraph{ACL2.}
Among white-box proof automation tactics, only ACL2's \enquote{waterfall} (TODO ref) has special support for forward reasoning.
(Some such tactics, e.g.\ Coq's auto, can execute arbitrary tactics including forward rules, but make no special provisions for them.)
It uses essentially Aesop's naive algorithm, i.e.\ the context of each goal is saturated independently without reusing partial results from other goals.
ACL2 additionally supports \emph{trigger terms}, which allow forward rules to be selected based on the presence of a particular term anywhere in the goal, rather than based on particular hypotheses.
Aesop implements a similar mechanism, \emph{rule patterns}, not described here due to space constraints.

@@

We have presented a technique that integrates forward reasoning into a tree-based proof search procedure supporting arbitrary proof rules.
We avoid redundant reasoning steps by caching partial forward rule applications in a data structure, the \emph{forward match state}, that enables efficient updates.
The technique is independent of the search strategy and logic used.
Our evaluation shows a substantial speedup on some synthetic benchmarks, relative to Aesop's previous implementation.
\end{document}

>>>>

% TODO: standardise indices vs indexes
% TODO: standardise dots after equations
% TODO: rename variable maps to variable indices everywhere
% TODO: Title Case for paragraphs

@@

In many proof assistants, simple \enquote*{white-box} automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's (TODO ref) and Rocq's (TODO ref) auto, ACL2's (TODO ref) waterfall, Lean's (TODO ref) Aesop (TODO ref), etc.
These tactics operate on \emph{goals} $Γ ⊢ A$, where the \emph{context} $Γ$ is a list of assumptions and $A$ is the proposition to be proved.
Faced with an initial goal, the tactics try various user-specified \emph{rules}, which correspond to admissible inferences in the prover's logic.
A successful rule reduces the current goal to zero or more subgoals, which are solved recursively until either a full proof is found or no rule applies to any open goal or the search times out.

@@

\paragraph{ACL2 and PVS.}
Among white-box proof automation tactics, to our knowledge only ACL2's \enquote{waterfall} (TODO ref) and PVS's various proof search strategies (TODO ref) have special support for forward reasoning.
Both use essentially Aesop's naive algorithm, i.e.\ the context of each goal is saturated independently without reusing partial results from other goals.
ACL2 additionally supports \emph{trigger terms}, which allow forward rules to be selected based on the presence of a particular term anywhere in the goal, rather than based on particular hypotheses.
Aesop implements a similar mechanism, \emph{rule patterns}, not described here due to space constraints.
Some other white-box tactics, e.g.\ Rocq's auto, can execute arbitrary tactics including forward rules, but make no special provisions for them.

@@

We have presented a technique that integrates forward reasoning into a tree-based proof search procedure supporting arbitrary proof rules.
We avoid redundant reasoning steps by caching partial forward rule applications in a data structure, the \emph{forward match state}, that enables efficient updates.
The technique is independent of the search strategy and logic used.
Our evaluation shows a substantial speedup on some synthetic benchmarks, relative to Aesop's previous implementation.

\end{document}
